\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-01 coqimpl.tex
Explaining sf
Adding more refering back to consistency issue.

These are some core extensions we proposed to support \cref{fig:STLC-example}. We use "Family" and "EndFamily" to scope, and "Field" is used to define new members in a family. "FieldClaim" is similar as "Field" but will invoke proof-mode in Coq and thus the user can use tactic programming to define new members.\footnote{We expect "FieldClaim" to be used in a \textit{proof-irrelevant} situation and thus whether overridable or not doesn't matter.}
We can define nested families.

@@

% explaining inheritance
Besides standalone family, we also need to implement family inheritance.
To simplify the implementation, we can consider all families as
inheritance and the standalone families are extending the empty family.
Thus, we need to deal with only inheritance during interactive theorem proving.

There are three kinds of inheritance---extend, override, and inherit. We
use first-order data to encode inheritance data inside the
implementation of the plugin. The inheritance data will include the
information of (1) the ``type'' of the parent family, (2) a list of
operation indicating how each field of the parent will be dealt with
(either inherited or overriden), (3) the operations indicating newly
extended fields, and (4) the ``type'' of the children family.
The surface syntax for the three kinds of inheritance
are demonstrated in the exemplar \cref*{fig:plugin-example3}. The
implementation of overriding, inheritance, and extension is achieved by
simply manipulating the internal representation of a family
correspondingly---we can swap the module in the list for overriding,
retain the module in the list for inheritance, and add new module into
the list for extension. The compilation from a family (internal
representation) to a module still acts the same.

Though inheritance is solely a first-order data rather
than a higher-order functor, it is still possible to
\text{mix-in} two inheritance data---by carefully ``mix'' the internal list, we can compose the inheritance. However, a good definition of mixin is still under investigation.\YZ{How about making mixin composition one of your Req's? Also does any example in the paper illustrate how to do this?}\EDJreply{mixin is not really formalized. There is no example illustrate it. The only example is the STLC example at the very end and I just kind of ``mention'' it. I think here I just want to emphasize, inheritance as first order data is still possible to support composition.}

@@

% explain ftheorem wrapping this complicated recursor
Finally, we also provide a command "FTheorem" in programming interface when the users want to use tactic
programming instead. This command (1) can avoid most boilerplate code (e.g., specify the handler family) when writing a recursor like above, (2) invoke
proof interaction mode and thus allow tactic programming, (3) and is also open to extension like "FRecursor".\YZ{example?}\EDJreply{FTheorem is really just a short hand for "FRecursor" + handler family. Nothing more conceptual here. If the reader is interested in the concrete syntax, I think I need to make the two examples avaliable in the appendix.}

@@

Here we show the example to prove "value_not_app" in
\cref{fig:plugin-example-global-reasoning-meta-data}, the script inside
curly brackets after "by" is the attached proof script that will be
executed upon "EndFamily", which triggers the compilation of the family
being defined.
% \YZ{Please check if my wording is correct.}\EDJreply{Yes. Resolved}
Since we are dealing with a concrete (vanilla) inductive type during compilation,
we can use Coq's "inversion" tactic to complete this proof.
This "Closing Fact", as an assertion, will have no problem being
inherited, but the proof script will be executed every time a derived
family is compiled.

>>>>

These are some core extensions we proposed to support \cref{fig:STLC-example}. We use "Family" and "EndFamily" to scope, and "Field" is used to define new members in a family. "FieldClaim" is similar as "Field" but will invoke proof-mode in Coq and thus the user can use tactic programming to define new members.\footnote{We expect "FieldClaim" to be used in a \textit{proof-irrelevant} situation and thus whether overridable or not doesn't matter.} Both "Field" and "FieldClaim" are only allowed to refer to the former fields, to avoid possible inconsistency mentioned in \ref{chg:consistency}.
We can define nested families.

@@

% explaining inheritance
\textbf{Implementing Inheritance.}
To resolve \ref{chg:consistency}, we make sure when inheritance happens, not only the content will be inherited, but the context of the corresponding field is also ``inherited''. Thus the mentioned informal example in \ref{chg:consistency} will not apply here because the overriding "f" is also in the context where "self" is empty and has no "g" in "self".

To simplify the implementation, we can consider all families as
inheritance and the standalone families are extending the empty family.
Thus, we need to deal with only inheritance during interactive theorem proving.

There are three kinds of inheritance---extend, override, and inherit. We
use a list to encode inheritance data inside the
implementation of the plugin. The inheritance data will include the
information of (1) the ``type'' of the parent family, (2) a list of
operation indicating how each field of the parent will be dealt with
(either inherited or overriden), (3) the operations indicating newly
extended fields, and (4) the ``type'' of the children family.
The surface syntax for the three kinds of inheritance
are demonstrated in the exemplar \cref*{fig:plugin-example3}. The
implementation of overriding, inheritance, and extension is achieved by
simply manipulating the internal representation of a family
correspondingly---we can swap the module in the list for overriding,
retain the module in the list for inheritance, and add new module into
the list for extension. The compilation from a family (internal
representation) to a module still acts the same.

Though inheritance is solely a first-order data rather
than a higher-order functor, it is still possible to
\text{mix-in} two inheritance data---by carefully ``mix'' the internal list, we can compose the inheritance. However, a good definition of mixin is still under investigation. We will show one practical example of \textit{mixin} later.\YZ{How about making mixin composition one of your Req's? Also does any example in the paper illustrate how to do this?}\EDJreply{mixin is not really formalized. There is no example illustrate it. The only example is the STLC example at the very end and I just kind of ``mention'' it. I think here I just want to emphasize, inheritance as first order data is still possible to support composition.}

@@

% explain ftheorem wrapping this complicated recursor
This design of decoupling of the case handlers and exhaustiveness checking handles the semantic well, however, it brings a lot of boilerplate code---we need to create an extra (handler) family, manually specifying the type of each case handlers, and then using "FRecursor" to ``tie the knot'' and construct a recursor field from the family.  
We thus provide a command "FTheorem" to remove these boilerplate in programming interface when using tactic programming instead. 
This command (1) can avoid specifying the handlers family and the type of each case handlers when writing induction like above, (2) invoke
proof interaction mode and thus allow tactic programming, (3) and is also open to extension like "FRecursor".\YZ{example?}\EDJreply{FTheorem is really just a short hand for "FRecursor" + handler family. Nothing more conceptual here. If the reader is interested in the concrete syntax, I think I need to make the two examples avaliable in the appendix.} We expect this "FTheorem" to be used in theorem proving but not general programming.

@@

Here we show the example to prove "value_not_app" in
\cref{fig:plugin-example-global-reasoning-meta-data}, the script inside
curly brackets after "by" is the attached proof script that will be
executed upon "EndFamily", which triggers the compilation of the family
being defined.
Since we are dealing with a concrete (vanilla) inductive type during compilation,
we can use Coq's "inversion" tactic to complete this proof.
This "Closing Fact", as an assertion, will have no problem being
inherited, but the proof script will be executed every time a derived
family is compiled.

### 2022-07-01 lang-design.tex
Save, fixing notation.
Adding more refering back to consistency issue.

\textlabel{Challenge (3)}{chg:consistency}~\textbf{Consistency vs. Self Parametrization}. The metatheoretical formulation in~\citet{zm2017} for family polymorphism is very natural as they consider extension of the family as the concatenation of a new linkage. However, this formulation is not directly adaptable for dependent type theory because it can easily encode non-termination (mutual calling of two function member) sabotaging consistency by the presence of "self". This conflict between self parametrization and consistency will be another challenge---we need change the original formulation of family and inheritance for dependent type theory.

@@

% we want to be able to reason pattern matching as well
% However, there is still difference between data recursion and induction
% reasoning, because the former one is \textit{proof-relevant}. This
% difference leads to another issue: \textlabel{Req (4)}{chg:definition-relevant-reasoning}:
% \textbf{we need to reason about (the computation about) extensible
% pattern matching}, just like "subst_lemma" in \cref{fig:STLC-example}.
% When proving "subst_lemma", we have to know "subst" is invariant on
% "tt", i.e., \mintinline{Coq}{(subst i x tt) = tt}. This kind of
% information requires exposing \textit{computational rules} from the
% recursors.
\YZ{Seems like what's missing here is to
convince the reader why exposing the definition of subst could be a
problem for family polymorphism.}\EDJreply{Actually I don't see why I need to emphasize this "as a problem". This is just a requirement like any other requirements earlier. I don't know how to show its difficulty at this early stage.}\YZreply{There is apparently a tension between the need to expose definitional equality and the ability to override. Looking at the bigger picture, I think this section reads better if phrased as discussing challenges posed by adding family polymorphism to a proof assistant. I have some initial ideas that I need to discuss with you.}

>>>>

\textlabel{Challenge (3)}{chg:consistency}~\textbf{Consistency vs. Self Parametrization}. The metatheoretical formulation in~\citet{zm2017} for family polymorphism is very natural as they consider extension of the family as the concatenation of a new linkage. However, this formulation is not directly adaptable for dependent type theory because it can easily encode non-termination sabotaging consistency by the presence of "self". Informally, in the orignal formulation, we can override the "f" in family $\{"f" := id; "g" := "call self.f"\}$ with "call self.g", resulting the mutual calling of "f" and "g" thus a deadloop. This conflict between self parametrization and consistency will be another challenge---we need change the original formulation of family and inheritance for dependent type theory.

@@

% we want to be able to reason pattern matching as well
% However, there is still difference between data recursion and induction
% reasoning, because the former one is \textit{proof-relevant}. This
% difference leads to another issue: \textlabel{Req (4)}{chg:definition-relevant-reasoning}:
% \textbf{we need to reason about (the computation about) extensible
% pattern matching}, just like "subst_lemma" in \cref{fig:STLC-example}.
% When proving "subst_lemma", we have to know "subst" is invariant on
% "tt", i.e., \mintinline{Coq}{(subst i x tt) = tt}. This kind of
% information requires exposing \textit{computational rules} from the
% recursors.
\YZ{Seems like what's missing here is to
convince the reader why exposing the definition of subst could be a
problem for family polymorphism.}\EDJreply{Actually I don't see why I need to emphasize this "as a problem". This is just a requirement like any other requirements earlier. I don't know how to show its difficulty at this early stage.}\YZreply{There is apparently a tension between the need to expose definitional equality and the ability to override. Looking at the bigger picture, I think this section reads better if phrased as discussing challenges posed by adding family polymorphism to a proof assistant. I have some initial ideas that I need to discuss with you.}\EDJreply{Design Space section is rewritten into Challenge now.}

### 2022-07-01 syntax.tex
Adding more refering back to consistency issue.

@@

Inheritance are
judgement so naturally second class citizen data (i.e. not something
function can return and cannot be assigned to variables), and it looks
very much like a function that transform a given linkage.

The syntax of the inheritance judgement is very close to how Family is
defined in the surface syntax---a Family with no parent is an
inheritance with empty input; an extended Family is exactly an
inheritance---because the parent of an extended Family can be overridden
thus inheritance should only define upon the ``interface'' instead of a
concrete family. However, since our current formulation doesn't support
further-binding (i.e. we cannot do inheritance inside a family), our
formulation still has some distance to the real family polymorphism.

>>>>

Now, together with sealing, we can read \ruleref{Sig Add} and \ruleref{Lnkg Add} out clearly: given a constructed linkage $o : \cL\sigma$ of signature $\sigma$, the next added field $t : T$ will have to be in the context of the abtraction $A$ of the compilation $\cC\sigma$ (witnessed by the seal $f$). The ``concatenation'' of the $o$ and $t$ leads to a new linkage, of which the signature is the ``concatenation'' of the $\sigma$ and $T$ (we also record the sealing $f$ for simplicity). The construction of linkage in \ruleref{Lnkg Add} requires $t$ to \textbf{only} refer to the (abstraction $A$ of) former part of the defining family $o : \cL\sigma$, this is to avoid inconsistency mentioned in \ref{chg:consistency} in metatheory.

@@

Inheritance are
judgement so naturally second class citizen data (i.e. not something
function can return and cannot be assigned to variables), and it looks
very much like a function that transform a given linkage. Like plugin implementation and \ruleref{Lnkg Add}, we resolve \ref{chg:consistency} by making sure corresponding fields are in corresponding context during inheritance, and an overriding term can only refer to fields before it.

The syntax of the inheritance judgement is very close to how Family is
defined in the surface syntax---a Family with no parent is an
inheritance with empty input; an extended Family is exactly an
inheritance---because the parent of an extended Family can be overridden
thus inheritance should only define upon the ``interface'' instead of a
concrete family. However, since our current formulation doesn't support
nested inheritance (i.e. we cannot do inheritance inside a family but only top level), our
formulation still has some distance to the real family polymorphism.
